---
## Winter 2022
## Project: Importation Risk
## Zahra Mohammadi
title: "R Notebook"
output: html_document 
---
## load libraries
```{r}
library(dplyr)
library(ggplot2); theme_set(theme_bw())
library(viridis)
library(tidyverse)
library(lubridate)
library("readxl")
library(reshape2)
library(patchwork)  
library(scales)
library(zoo)
library(viridis)
library(RColorBrewer)
library(mapcan)


cpalete = c( "#999999", "#E69F00", '#336699',"#99FF33", "#CC6699", "#cc0000", "#414487FF" ,"#35B779FF", "#663399") 

```
## Read data 
```{r}
## 1. Statistics Canada (Frontier Counts (FC))
## cleaned and aggregated version focused on international port of entries in NL
## Data source detail:  Table 24-10-0041-01, DOI: https://doi.org/10.25318/2410004101-eng ##https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&SDDS=5005 

FC <- read.csv("Data/TraVolData/FC-NL.csv")
colnames(FC)[colnames(FC) == "REF_DATE"] <- "date"
FC$date <- as.Date(as.character(FC$date), format = "%Y-%m-%d")
FC <- FC[FC$date < as.Date("2021-06-01"),] 

FCEXCrew <- read.csv("Data/TraVolData/FC-NL-nocrews.csv")
colnames(FCEXCrew)[colnames(FCEXCrew) == "REF_DATE"] <- "date"
FCEXCrew$date <- as.Date(as.character(FCEXCrew$date), format = "%Y-%m-%d")
FCEXCrew <- FCEXCrew[FCEXCrew$date < as.Date("2021-06-01"),] 

## 2. IATA (just inbound travel volume that is extracted form IATA data)
IATA <- read.csv("Data/TraVolData/InboundIATA.csv")
IATA$date <- as.Date(as.character(IATA$date), format = "%Y-%m-%d")

## 3. TDF ( aggregated )
TDF <- read.csv("Data/TraVolData/ExtractTDF.csv")
TDF$date <- as.Date(as.character(TDF$date), format = "%Y-%m-%d") 

## 4. Reported travel related cases in NL
Travel.case <- read.csv("Data/Travel.Case.csv")
Travel.case$date <- as.Date(as.character(Travel.case$date), format = "%Y-%m-%d")
```
##plot TDF data
```{r}
# get monthly per province from categorical regression (plot)
r.TDF <- TDF
##add month and year column
r.TDF$year  <- strftime(r.TDF$date, "%Y")   # Create year column
r.TDF$month <- strftime(r.TDF$date, "%m")   # Create month column


lm.TDF <- function(){
  model_ON <- lm(ON ~ month + year , data = r.TDF)      
  model_PEI <- lm(PEI ~ month + year , data = r.TDF)  
  model_QC <- lm(QC ~ month + year , data = r.TDF)      
  model_SK <- lm(SK ~ month + year , data = r.TDF)  
  model_AB <- lm(AB ~ month + year , data = r.TDF)      
  model_BC <- lm(BC ~ month + year , data = r.TDF)  
  model_MB <- lm(MB ~ month + year , data = r.TDF)      
  model_NB <- lm(NB ~ month + year , data = r.TDF)  
  model_NS <- lm(NS ~ month + year , data = r.TDF)      
  model_TR <- lm(TR ~ month + year , data = r.TDF) 
  model_US <- lm(US ~ month + year , data = r.TDF)      
  model_nonUS <- lm(nonUS ~ month + year , data = r.TDF) 
  model_ALLCANotNL <- lm(ALLCANotNL ~ month + year , data = r.TDF)  
  model_AllFR <- lm(AllFR ~ month + year , data = r.TDF)  

  r.TDF$ON <- model_ON[["fitted.values"]]
  r.TDF$PEI <- model_PEI[["fitted.values"]]
  r.TDF$QC <- model_QC[["fitted.values"]]
  r.TDF$SK <- model_SK[["fitted.values"]]
  r.TDF$AB <- model_AB[["fitted.values"]]
  r.TDF$BC <- model_BC[["fitted.values"]]
  r.TDF$MB <- model_MB[["fitted.values"]]
  r.TDF$NB <- model_NB[["fitted.values"]]
  r.TDF$NS <- model_NS[["fitted.values"]]
  r.TDF$TR <- model_TR[["fitted.values"]]
  r.TDF$nonUS <- model_nonUS[["fitted.values"]]
  r.TDF$ALLCANotNL  <- model_ALLCANotNL[["fitted.values"]]
  r.TDF$AllFR <- model_AllFR[["fitted.values"]]

  return(r.TDF)
}
#regression and cleanup
r.TDF = lm.TDF()
 
# aggregating the daily data
mon.TDF <- r.TDF %>%                        
  group_by( year, month) %>% 
  dplyr::summarize(ON = sum(ON) , PEI = sum(PEI) , QC = sum(QC),
                   SK = sum(SK), AB = sum(AB), BC = sum(BC),
                   MB = sum(MB), NB = sum(NB), NS = sum(NS),
                   TR= sum(TR) ,
                   US = sum(US), nonUS = sum(nonUS),
                   ALLCANotNL = sum(ALLCANotNL), AllFR=sum(AllFR)) %>%
  as.data.frame()


#make one column for date
mon.TDF <- mon.TDF %>%
  mutate(date = with(., sprintf("%s-%02s", year, month)))
mon.TDF$date <- as.Date(paste(mon.TDF$date,"-01",sep=""))  # convert to date
mon.TDF <- mon.TDF  %>% relocate(date, .before = year) # organising
mon.TDF <- select(mon.TDF , -c(year, month))

# plot
colLeg <- c("INT" = cpalete[1], "CA" = cpalete[2])
p1 <- ggplot() + 
  geom_point(data=TDF, aes(x=date, y= ALLCANotNL, color='CA')) +
  geom_line( data=r.TDF, aes(x=date, y= ALLCANotNL, col='CA'), size=1) +
  geom_point(data=TDF, aes(x=date, y= AllFR, color= 'INT')) +
  geom_line( data=r.TDF, aes(x=date, y=AllFR, col='INT'), size=1) +
  
  labs(x="", y="\n Arriving travelers (daily)", title="Travel volume (TDF)") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  scale_color_manual(values = colLeg) +
  guides(color = guide_legend(title = "Departure Origin"), size= rel(1.4))+
  theme_classic() + theme(axis.text.x = element_text(angle = 90, size=rel(1.2)), 
                          #legend.title = title('Departure Origin'),
                          legend.position="bottom",
                          legend.text=element_text(size=rel(1.2)),
                          plot.title=element_text(size=rel(1.3), face="bold"),
                          axis.title = element_text(size=rel(1.1)),
                          axis.text.y = element_text(size=rel(1.2)))


p1
```

##plot Travel-related cases
```{r}
ave.fun <- function(){
  #add month-date
  Travel.case$year  <- strftime(Travel.case$date, "%Y")   
  Travel.case$month <- strftime(Travel.case$date, "%m") 
  Ave.df1 <- aggregate(Tot.Int ~ month + year , Travel.case , mean )
  Ave.df2 <- aggregate(Tot.Dom ~ month + year , Travel.case , mean )
  Ave.df <- Ave.df1 %>% left_join(Ave.df2 , by= "month")
  Ave.df  <- Ave.df  %>% mutate(date = with(., sprintf("%s-%02s", year.x, month)))
  Ave.df$date <- as.Date(paste(Ave.df$date,"-01",sep=""))  # convert to date
  Ave.df <- select(Ave.df  , c(date, Tot.Int, Tot.Dom))
  return( Ave.df )
}

mon.Travel.case <- ave.fun()

## repeat average for days in a month for plotting purpose 
rep.fun <- function(mon.Travel.case){
  df <- mon.Travel.case
  df["Ym"] <- format(as.Date(df$date), "%Y-%m")
  Av_daily <- df %>% select(date) %>% 
    complete(date=  seq.Date(min(date), max(date)+30, by="day"),) %>%
    mutate(Ym=paste(year(date),str_pad(month(date), 2, pad = "0"),sep = "-"))
  Av_daily <- Av_daily %>% left_join(df, by="Ym")
  Av_daily <- select(Av_daily, c("date.x", "Tot.Int" ,"Tot.Dom" ))
  Av_daily <- Av_daily %>% rename(date = date.x)
  return(Av_daily)
}

df.plt <- rep.fun(mon.Travel.case)
df.plt <- select(Travel.case, c("date", "Tot.Int", "Tot.Dom" )) %>% left_join(df.plt , by="date")
df.plt[df.plt == 0] <- NA

colLeg <- c("INT" = cpalete[1], "CA" = cpalete[2])
p2 <- ggplot() + 
  geom_point(data=df.plt, aes(x=date, y= Tot.Dom.x, color='CA')) +
  geom_line( data=df.plt, aes(x=date, y= Tot.Dom.y, col='CA'), size=1) +
  geom_point(data=df.plt, aes(x=date, y= Tot.Int.x, color= 'INT')) +
  geom_line( data=df.plt, aes(x=date, y=Tot.Int.y, col='INT'), size=1) +
  
  labs(x="", y="\n Infected travelers (daily)", title="Travel-related cases (NLHS)") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  scale_color_manual(values = colLeg) +
  guides(color = guide_legend(title = "Departure Origin"), size= rel(1.3))+
  theme_classic() + theme(axis.text.x = element_text(angle = 90, size=rel(1.2)),
                          legend.position="bottom",
                          legend.text=element_text(size=rel(1.2)),
                          plot.title=element_text(size=rel(1.3), face="bold"),
                          axis.title = element_text(size=rel(1.1)),
                          axis.text.y = element_text(size=rel(1.2)))

p2
```
##Visualize pre pandemic data sources gap
```{r}
## show differences in international travelers from IATA nd FC
df.plt <- select(IATA, c("date", "ALLCANotNL", "AllFR"))
df.plt <- df.plt %>% left_join(FC, by='date') 

colLeg <- c("IATA" = cpalete[6], "FC" = cpalete[9])
p3 <- ggplot() + 
  geom_point( data= df.plt, aes(x=date, y=AllFR ,  color='IATA')) +
  geom_point( data= df.plt, aes(x=date, y=total , color='FC')) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  labs(x="", y="\n Arrival travelers", title="Pre-pandemic, International origin") +
  scale_color_manual(values = colLeg) +
  guides(color = guide_legend(title = "Data Source"), size=rel(1.4)) +
  theme_classic() + theme(axis.text.x = element_text(angle = 90, size=rel(1.3)), 
                          legend.position="bottom",
                          legend.text=element_text(size=rel(1.2)),
                          plot.title=element_text(size=rel(1.4), face="bold"),
                          axis.title = element_text(size=rel(1.3)),
                          axis.text.y = element_text(size=rel(1.3)))
p3
```
##visualize during pandemic data sources gap
```{r}
## show differences in international travelers from TDF nd FC
df.plt <- select(mon.TDF, c("date", "ALLCANotNL", "AllFR")) %>% left_join(FC, by ="date")

colLeg <- c("TDF-INT" = cpalete[6], "FC" =cpalete[9]) 
p4 <- ggplot() + 
  geom_point( data= df.plt, aes(x=date, y=AllFR ,  color='TDF-INT')) +
  geom_point( data= df.plt, aes(x=date, y=total , color='FC')) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  labs(x="", y="\n Arrival travelers", title="During pandemic, International origin") +
  scale_color_manual(values = colLeg) +
  guides(color = guide_legend(title = "Data Source"), size= rel(1.4))+
  theme_classic() + theme(axis.text.x = element_text(angle = 90, size=rel(1.3)), 
                          legend.position="bottom",
                          legend.text=element_text(size=rel(1.2)),
                          plot.title=element_text(size=rel(1.4), face="bold"),
                          axis.title = element_text(size=rel(1.3)),
                          axis.text.y = element_text(size=rel(1.3)))


p4
```

## get the total travel volume - pre pandemic
```{r}
total.pre <- function(IATA, FC, FCEXCrew) {
  date <- seq(as.Date("2019-01-01"),as.Date("2020-03-01"),by = "month")  
  df <- as.data.frame(date)
  df["Air_CA"] <- IATA$ALLCANotNL[1:15] 
  df["CA1"] <- NA
  #April-August 
  for (i in 4:8){ df$CA1[i] <- (df$Air_CA[i])*(1/0.73)  }
 # Jan, Feb 2019/2020, Dec
  for (i in 1:2){df$CA1[i] <- (df$Air_CA[i])*(1/0.88)}
  for (i in 12:14){df$CA1[i] <- (df$Air_CA[i])*(1/0.88)}
  # sep- Nov, March2019/2020
  for (i in 9:11){df$CA1[i] <- (df$Air_CA[i])*(1/0.81)}
  df$CA1[3] <- (df$Air_CA[3])*(1/0.81)
  df$CA1[15] <- (df$Air_CA[15])*(1/0.81)
  #crews members from Int
  df["crew_Int"] <- FC$total[1:15] - FCEXCrew$total[1:15]
  #portion of travelers as crew members from Canadian origins(# Assume twice crew_int)
  df["CA"] <- df$CA1 + 2.5*df$crew_Int
  df["CA"] <- round(df["CA"]) 
  df["Int"] <- FC$total[1:15]
  
  df <- select(df, c("date" ,"CA", "Int"))
  return(df)
}

TRV.Pre <- total.pre(IATA, FC, FCEXCrew)

```
## get the total travel volume - during the pandemic
```{r}
total.pand <- function(TDF, FC, FCEXCrew){
  df <- FC %>% filter(FC$date > as.Date("2020-02-01") )
  colnames(df)[colnames(df) == "total"] <- "FC"
  ##add FCEXCrew
  df <- df %>% left_join( FCEXCrew, by='date')
  colnames(df)[colnames(df) == "total"] <- "FC_Excrew"
  df["crew_Int"] <- df$FC - df$FC_Excrew
  df2 <- select(mon.TDF, c("date" , "ALLCANotNL", "AllFR" )) %>% 
    rename( TDF_CA = ALLCANotNL , TDF_Int = AllFR)
  df <- df2 %>% left_join( df, by='date')
  #Exempt from International,  compare (FCInt) vs NLCHI_Int (should include FCNonRes!)
  df["Exp_Int"] <-  df$FC_Excrew -  df$TDF_Int 
  #Exempt from CA
  df["crew_CA"] <- 2.5*df$crew_Int  #crew from CA
  df["Exp_CA"] <- 2.4*df$Exp_Int    #excluded crew
  # update CA 
  df["CA"] <- df$TDF_CA + df$Exp_CA + df$crew_CA
  df["Int"] <- df$FC
  df["CA"] <- round(df$CA)
  
  #df <- select(df, c("date" ,"CA", "Int"))
  return(df)
}

TRV.Pan <- total.pand(TDF, FC, FCEXCrew)
TRV.Pan <- select(TRV.Pan, c("date" ,"CA", "Int"))
```
## merging pre and during pandemic and make one dataframe of all total travel volume 
```{r}
Final_data <- rbind(TRV.Pre, TRV.Pan )
Final_data["estimated"] <- rowSums(Final_data[,2:3], na.rm=TRUE)
```
## plot all time (estimated total travel volume)
```{r}
df.plt <- select(Final_data,  c("date","Int" ,"CA" ))
df.plt<- df.plt%>% rename( INT = Int )
df.plt <- subset(df.plt, !date %in% c(as.Date("2019-01-01"), as.Date("2019-02-01")) ) #dont plot Jan&Feb20119

df.plt <- melt(df.plt, id=c("date"))

cut1 <- data.frame(Ref = c("Public Health Emergency in NL"),
                   vals = c(as.Date("2020-03-20")),
                   stringsAsFactors = FALSE)

cut2 <- data.frame(Ref = c( "Travel Declaration Form data begins"),
                   vals = c(as.Date("2020-08-15")),
                   stringsAsFactors = FALSE)


p5 <- ggplot() +
  geom_bar(data=df.plt , aes(x=date, y=value, fill=variable), stat="identity") +
  geom_vline(mapping = aes(xintercept = vals), linetype="dashed", data = cut1, 
             show.legend = FALSE, size=0.3) +
  geom_vline(mapping = aes(xintercept = vals), linetype="dashed", data = cut2, 
             show.legend = FALSE, size=0.3) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  labs(color=" ", 
       title="Total travel volume",
       y="\n Arriving travelers (monthly)",
       x="") +
  guides(fill = guide_legend(title = "Departure Origin"))+
  theme(legend.position="right") +
  scale_fill_manual(values=c(cpalete[1], cpalete[2])) + 
  geom_text(mapping = aes(x = vals, y = 210000, label = Ref, hjust = 1.2, vjust =  1.5, angle = 90),
           data = cut1, size=rel(3.1))  +
  geom_text(mapping = aes(x = vals, y = 220000, label = Ref, hjust = 1.1, vjust = - 0.5, angle = 90),
           data = cut2, size=rel(3.1))  +
  theme_classic() + theme(axis.text.x = element_text(angle = 90, size=rel(1.2)), 
                          legend.position="bottom" ,
                          legend.text=element_text(size=rel(1.2)),
                          plot.title=element_text(size=rel(1.3), face="bold"),
                          axis.title = element_text(size=rel(1.1)),
                          axis.text.y = element_text(size=rel(1.2)))

p5

```

##plot reduction (estimated total travel volume and raw data)
```{r}
df.plt <- FC
df.plt <- df.plt %>% rename(FC = total)
df.plt <- df.plt %>% left_join( select(mon.TDF, c("date", "ALLCANotNL", "AllFR")), by='date')
df.plt["TDF"] <- df.plt$ALLCANotNL + df.plt$AllFR
df.plt <- df.plt  %>% left_join(select(IATA, c("date", "ALLCANotNL", "AllFR")), by='date') 
df.plt["IATA"] <- df.plt$AllFR.y + df.plt$ALLCANotNL.y
df.plt <- df.plt %>% left_join(Final_data, by= 'date') 
df.plt <- select(df.plt , c("date", "FC", "TDF", "IATA" , "estimated"))

# remove Jan and Feb 2019
df.plt <- subset(df.plt, !date %in% c(as.Date("2019-01-01"), as.Date("2019-02-01")) )


cuts <- data.frame(Ref = c("1", "2", "3", "4", "5", "6","7","8","9"),
                   vals = c(as.Date("2020-03-13"), as.Date("2020-05-04"),as.Date("2020-06-09"),
                            as.Date("2020-07-03"), as.Date("2020-08-31"), as.Date("2020-10-20"), 
                            as.Date("2020-11-25"),as.Date("2021-02-01"),
                            as.Date("2021-03-27")), stringsAsFactors = FALSE)

cut3 <- data.frame(Ref = c("81.6% decrease"),
                   vals = c(as.Date("2021-01-28")),
                   stringsAsFactors = FALSE)


colLeg <- c("IATA" = cpalete[4], "TDF" = cpalete[5],
            "FC" = cpalete[3], "Estimated"= cpalete[6])

p6 <- ggplot() +
  geom_line(data=df.plt, aes(x=date, y= estimated, col="Estimated"), lty=1, size=1)+
  geom_line(data=df.plt, aes(x=date, y= IATA, col="IATA"), lty=1, size=0.7)+
  geom_line(data=df.plt, aes(x=date, y= TDF, col="TDF"), lty=1, size=0.7)+
  geom_line(data=df.plt, aes(x=date, y= FC, col="FC"), linetype="dashed", lty=2, size=0.7)+
  annotate("rect", xmin = as.Date("2020-03-01"), xmax =as.Date("2020-09-01"), ymin = 0,
           ymax = 210000, alpha = .1)+
  geom_vline(mapping = aes(xintercept = vals), linetype="dashed", data = cuts, 
             show.legend = FALSE, size=0.3) +
  geom_text(mapping = aes(x = vals, y = 200000, label = Ref, hjust = 1.2, vjust = -.5), 
            data = cuts, size=3, fontface="bold") +
  geom_text(mapping = aes(x = vals, y = 30000, label = Ref, hjust = .13, vjust =  0.1, angle = 0),
           data = cut3, size=rel(3.2), color= cpalete[6], fontface="bold")  +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  theme_light() + labs(color=" ", title="Travel volume", y="\n Arriving travelers (monthly)", x="") +
  scale_color_manual(values = colLeg) +
  theme_classic() + theme(axis.text.x = element_text(angle = 90, size=rel(1.2)), 
                          legend.title = element_blank(),
                          legend.position="bottom",
                          legend.text=element_text(size=rel(1.2)),
                          plot.title=element_text(size=rel(1.3), face="bold"),
                          axis.title = element_text(size=rel(1.1)),
                          axis.text.y = element_text(size=rel(1.2)))

p6

```

## calculate reduction
```{r}
## March2019-Feb2020 (not June-july-August since during pandemic we dont have thoes)
reduct <- function(Final_data){
  df <- Final_data
  df <- subset(df, !date %in% c(as.Date("2019-01-01"), as.Date("2019-02-01"), as.Date("2020-03-01"),
                              as.Date("2019-06-01"), as.Date("2019-07-01"), as.Date("2019-08-01")) )
  rownames(df) <- NULL  # Reset row names
  tot20 <- sum(df[1:9, 'estimated'], na.rm = TRUE)  #2019-2020 (before pan)
  tot21 <- sum(df[10:18, 'estimated'], na.rm = TRUE)   #2020-2021 (pan)
  ##percentage decrease
  decrease <- tot20 - tot21
  prdecrease <- (decrease/tot20)*100
  return(prdecrease)
}

reduct(Final_data) 
```

## merging Final volume with percentage (NLCHI -IATA) - corrected - to find share of each provinces
```{r}
per12  <- read.csv("Data/TraVolData/percentage(h).csv")
per12$date <- as.Date(as.character(per12$date), format = "%Y-%m-%d")

ProFinal <- per12 %>% left_join(Final_data, by= 'date')

#make volume for origin/month
for (i in 1:nrow(ProFinal)){
  for (j in 2:11){
     ProFinal[i,j] <- (ProFinal[i,j]*ProFinal$CA[i])/100
  }
  for (j in 12:13){
     ProFinal[i,j] <- (ProFinal[i,j]*ProFinal$Int[i])/100
  }
}

#percentage (share of each province monthly) 
CRPer <- ProFinal
for (i in 1:nrow(CRPer)){
  for (j in 2:length(CRPer)){
     CRPer[i,j] <- (CRPer[i,j]*100)/(CRPer$estimated[i])
  }
}

```
# plot percent after correction (provinces)
```{r}
#delete Jan and Feb 2019 from IATA 
plt.df <- select(CRPer, -c("Int", "CA", "estimated"))
plt.df <- subset(plt.df, !date %in% c(as.Date("2019-01-01"), as.Date("2019-02-01")) )

plt.df <- melt(plt.df, id=c("date"))

mycolors = viridis(12) 
colormap <- c("US" = mycolors[12], "nonUS" = mycolors[11], "ON" = mycolors[10],
              "NS" = mycolors[9], "AB" = mycolors[8], "BC" = mycolors[7], 
              "QC" = mycolors[6], "NB" = mycolors[5], "MB" = mycolors[4], 
              "SK" = mycolors[3], "PEI"= mycolors[2], "TR" = mycolors[1]
             )

cut1 <- data.frame(Ref = c("Public Health Emergency in NL"),
                   vals = c(as.Date("2020-03-20")),
                   stringsAsFactors = FALSE)

cut2 <- data.frame(Ref = c( "Travel Declaration Form data begins"),
                   vals = c(as.Date("2020-08-15")),
                   stringsAsFactors = FALSE)


p7 <- ggplot() +
  geom_bar(data=plt.df , aes(x=date, y=value, fill=variable),stat="identity") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %y") +
  geom_vline(mapping = aes(xintercept = vals), linetype="dashed", data = cut1, 
             show.legend = FALSE, size=0.3) +
   geom_vline(mapping = aes(xintercept = vals), linetype="dashed", data = cut2,
              show.legend = FALSE, size=0.3) +
  labs(color=" ", 
       title="Percentage of travelers arriving from different origins",
       y="\n ",
       x="") +
  scale_fill_manual(values = colormap) +
  guides(fill = guide_legend(title = "Departure Origin"), size= rel(1.4))+
  geom_text(mapping = aes(x = vals, y = 100, label = Ref, hjust = 1.2, vjust =  1.5, angle = 90),
           data = cut1, size=rel(3.1))  +
  geom_text(mapping = aes(x = vals, y = 100, label = Ref, hjust = 1.04, vjust = - 0.5, angle = 90),
           data = cut2, size=rel(3.1))  +
  theme_classic() + theme(axis.text.x = element_text(angle = 90, size=rel(1.2)), 
                          legend.position="bottom" ,
                          legend.text=element_text(size=rel(1.2)),
                          plot.title=element_text(size=rel(1.3), face="bold"),
                          axis.title = element_text(size=rel(1.1)),
                          axis.text.y = element_text(size=rel(1.2))) + 
  scale_y_continuous(labels = function(x) paste0(x*1, "%"))

p7

```

## Canada Map (pre and during the pandemic), show average percentage over 9 month
```{r}
Av.per <- function(CRPer){
  AA <- CRPer
  # get the average months March 2019 to Feb 2020 (except June July August 2019) 
  AA <- subset(AA, !date %in% c(as.Date("2019-01-01"), as.Date("2019-02-01"), 
                                as.Date("2020-03-01"), as.Date("2019-06-01"), 
                                as.Date("2019-07-01"), as.Date("2019-07-01"), 
                                as.Date("2019-08-01")) )
  rownames(AA) <- NULL   # Reset row names             
  AA1 <- AA[1:9,]        #pre-pandemic
  AA2 <- AA[10:18,]      #pandemic
  # get the mean for both dataframe
  A1 <- colMeans(AA1[2:13])
  A2 <- colMeans(AA2[2:13])
  
  A1 <- data.frame(A1)
  A1 <- cbind(province = rownames(A1), A1)
  rownames(A1) <- 1:nrow(A1)
  #repeat TR for three tritories
  A1[nrow(A1) + 1,] <- c( "YT",  0.338555269656022)
  A1[nrow(A1) + 1,] <- c( "NT", 0.338555269656022)
  A1[nrow(A1) + 1,] <- c( "NU",0.338555269656022)
  A1$A1 <- as.numeric(A1$A1)
  A1 <- A1 %>% rename(Percentage = A1)
  A1$province[A1$province == 'PEI'] <- 'PE'

  A2 <- data.frame(A2)
  A2 <- cbind(province = rownames(A2), A2)
  rownames(A2) <- 1:nrow(A2)
  #repeat TR for three tritories
  A2[nrow(A2) + 1,] <- c( "YT", 3.62752706114931)
  A2[nrow(A2) + 1,] <- c( "NT", 3.62752706114931)
  A2[nrow(A2) + 1,] <- c( "NU", 3.62752706114931)
  A2$A2 <- as.numeric(A2$A2)
  A2 <- A2 %>% rename(Percentage = A2)
  A2$province[A2$province == 'PEI'] <- 'PE'

  #return(A1)
  return(A2)
}

A1 <- Av.per(CRPer)
A2 <- Av.per(CRPer)


## canada map
pr_geographic <- mapcan(boundaries = province, type = standard)
pr_geographic <- inner_join(pr_geographic, A1, by = c("pr_alpha" = "province")) 

m1 <- pr_geographic %>%
  ggplot(aes(x = long, y = lat, group = group, fill = Percentage)) +
  geom_polygon(alpha= 1, show.legend = NA) +
  annotate(geom = "text", x = 642412.6, y = 201110.2, label = "27%", color = "white", size = 3)+
  annotate(geom = "text", x = 1349519.2, y = 782762.2,label = "14%", color = "white", size = 3)+
  annotate(geom = "text", x = -115306.8, y = 644233.4,label = "0.7%",color = "white", size = 3)+
  annotate(geom = "text", x = 2163606.1, y = 237946.6, label = "3%", color = "white", size = 3)+
  annotate(geom = "text", x = -1092152.6,y = 809288.2, label = "7%", color = "white", size = 3)+
  annotate(geom = "text", x = -1867594.7,y =912726.5, label = "2.5%",color = "white", size = 3)+
  annotate(geom = "text", x = -675361.6, y = 487741.4,label = "0.5%",color = "white", size = 3)+
  annotate(geom = "text", x = -888417.3, y = 1651194.8,label = "0.3%",color ="white", size = 3)+
  annotate(geom = "text", x = 2413606.1, y = 421394.6,label = "0.6%",color ="black", size = 3)+
  annotate(geom = "text", x = 2695757.5, y = 232727.5,label = "22%", color = "black", size = 3)+
  annotate(geom = "text", x = 1565757, y = -750409 , label = "US 14% & Others 8%", 
           fontface = "italic", color = "black", size = 3) +
  coord_fixed() +
  theme_mapcan() +
  scale_fill_continuous(type = "viridis", 
    limits = c(0,30), breaks = c(0, 5, 10, 15, 20, 25,30),
    guide = guide_colourbar(nbin = 100, draw.ulim = FALSE, draw.llim = FALSE,barwidth = 20,
                            title = "Travel volume (%)")
   )+
  theme_mapcan() + theme(legend.position="bottom",
                         plot.title=element_text(size=rel(1.3), face="bold"))+ 
  ggtitle("Before the pandemic")



pr_geographic2 <- mapcan(boundaries = province, type = standard)
pr_geographic2 <- inner_join(pr_geographic2, A2, by = c("pr_alpha" = "province")) 
m2 <- pr_geographic2 %>%
  ggplot(aes(x = long, y = lat, group = group, fill = Percentage)) +
  geom_polygon(show.legend = NA) +
  annotate(geom = "text",x = 642412.6,y = 201110.2, label = "21%", color = "white",size = 3)+
  annotate(geom = "text",x = 1349519.2,y = 782762.2, label = "4.7%",color = "white",size = 3)+
  annotate(geom = "text",x = -115306.8,y = 644233.4, label = "1%", color = "white",size = 3)+
  annotate(geom = "text",x = 2163606.1,y = 237946.6, label = "3%", color = "white", size = 3)+
  annotate(geom = "text",x = -1092152.6,y = 809288.2, label = "16%", color = "white",size = 3)+
  annotate(geom = "text",x = -1867594.7,y =912726.5, label = "5.2%", color = "white",size = 3)+
  annotate(geom = "text",x = -675361.6,y = 487741.4, label = "0.7%", color = "white",size = 3)+
  annotate(geom = "text",x = -888417.3,y = 1651194.8 , label = "3.6%", color ="white",size = 3)+
  annotate(geom = "text", x = 2413606.1, y = 421394.6,label = "0.7%",color ="black", size = 3)+
  annotate(geom = "text",x = 2695757.5,y = 232727.5, label = "19%", color = "black", size = 3)+
  annotate(geom = "text", x = 1565757, y = -750409 , label = "US 9% & Others 15%", 
           fontface = "italic", color = "black", size = 3) +
  coord_fixed() +
  scale_fill_continuous( type = "viridis",
    limits = c(0,30), breaks = c(0, 5, 10, 15, 20, 25,30),
    guide = guide_colourbar(nbin = 100, draw.ulim = FALSE, draw.llim = FALSE,
                            title = "Travel volume (%)")
    ) +
  theme_mapcan() + theme(legend.position="", 
                         plot.title=element_text(size=rel(1.3), face="bold"))+
  ggtitle("During the pandemic")


m1 + m2  
```
### Canada Map (pre and during the pandemic), show average daily over 9 month
```{r}
Av.daily <- function(ProFinal){
  CC <- ProFinal
  # get the average months March 2019 to Feb 2020 (except June July August 2019) 
  CC <- subset(CC, !date %in% c(as.Date("2019-01-01"), as.Date("2019-02-01"),
                                as.Date("2020-03-01"), as.Date("2019-06-01"), 
                                as.Date("2019-07-01"), as.Date("2019-07-01"), 
                                as.Date("2019-08-01")) )
  rownames(CC) <- NULL  # Reset row names              
  CC1 <- CC[1:9,]    #prepandemic
  CC2 <- CC[10:18,]  #pandemic
  # get the mean for both dataframe
  C1 <- colSums(CC1[2:13])
  C2 <- colSums(CC2[2:13])

  C1 <- data.frame(C1)
  C1 <- cbind(province = rownames(C1), C1)
  rownames(C1) <- 1:nrow(C1)
  #repeat TR for three tritories
  C1[nrow(C1) + 1,] <- c( "YT",  2682.378)
  C1[nrow(C1) + 1,] <- c( "NT", 2682.378)
  C1[nrow(C1) + 1,] <- c( "NU",2682.378)
  C1$C1 <- as.numeric(C1$C1)
  C1 <- C1 %>% rename(Daily = C1)
  C1$province[C1$province == 'PEI'] <- 'PE'
  # 273 days pre pandemic 2019
  C1$Daily <- (C1$Daily)/273
  C1$Daily <- ceiling(C1$Daily)


  C2 <- data.frame(C2)
  C2 <- cbind(province = rownames(C2), C2)
  rownames(C2) <- 1:nrow(C2)
  #repeat TR for three tritories
  C2[nrow(C2) + 1,] <- c( "YT", 5329)
  C2[nrow(C2) + 1,] <- c( "NT", 5329)
  C2[nrow(C2) + 1,] <- c( "NU", 5329)
  C2$C2 <- as.numeric(C2$C2)
  C2 <- C2 %>% rename(Daily = C2)
  C2$province[C2$province == 'PEI'] <- 'PE'
  ## 274 days during the pandemic 2020
  C2$Daily <- (C2$Daily)/274
  C2$Daily <- ceiling(C2$Daily)
  
  #return(C1)
  return(C2)
}

C1 <- Av.daily(ProFinal)
C2 <- Av.daily(ProFinal)

pr_geographic <- mapcan(boundaries = province, type = standard)
pr_geographic <- inner_join(pr_geographic, C1, by = c("pr_alpha" = "province")) 
df.label <- read.csv("Data/df.label2.csv")
df.label <- df.label %>% left_join(C1 , by= "province")
df.label$Daily <- round(df.label$Daily)


m3 <- pr_geographic %>%
  ggplot(aes(x = long, y = lat, group = group, fill = Daily)) +
  geom_polygon(show.legend = NA) +
  geom_text(data =df.label, aes(x = new.long, y = new.lat, group = 1, label = Daily), 
            size = 4, color = "white")+ 
  annotate(geom = "text",x = 2685757.5,y = 232727.5,label = "648", color = "black", size = 4)+
  annotate(geom = "text", x = 2413606.1, y = 421394.6,label = "19",color ="black", size = 4)+
  annotate(geom = "text", x = 1565757, y = -750409 , label = "US 479 & Others 300", 
           fontface = "italic", color = "black", size = 4) +
  coord_fixed() +
  theme_mapcan() +
  scale_fill_continuous(type = "viridis",
    limits = c(0,900), breaks = c(25,50,75, 100, 200, 300, 400, 500,600, 700,800,900),
    #limits = c(0,900), breaks = c(25,50,75,100,250,500,750,900),
    guide = guide_colourbar(nbin = 13, draw.ulim = FALSE, draw.llim = FALSE, barwidth = 20,
                           title = "Total travel volume (Daily)", label.hjust = 0.7, 
                           label.vjust = 0.79,
                           label.theme = element_text(angle = 90, size=7.9))
    ) +
  theme_mapcan() + theme(legend.position="bottom",
                         plot.title=element_text(size=rel(1.3), face="bold"))+
  ggtitle("Before the pandemic")


pr_geographic2 <- mapcan(boundaries = province, type = standard)
pr_geographic2 <- inner_join(pr_geographic2, C2, by = c("pr_alpha" = "province")) 
df.label <- read.csv("Data/df.label2.csv")
df.label <- df.label %>% left_join(C2 , by= "province")

m4 <- pr_geographic2 %>%
  ggplot(aes(x = long, y = lat, group = group, fill = Daily)) +
  geom_polygon(show.legend = NA) +
  geom_text(data =df.label, aes(x = new.long, y = new.lat, group = 1, label = Daily),
            size = 4, color = "white")+ 
  annotate(geom = "text", x = 2715757.5,y = 232727.5,label = "115", color = "black", size = 4)+
  annotate(geom = "text", x = 2413606.1, y = 421394.6,label = "5",color ="black", size = 4)+
  annotate(geom = "text", x = 1565757, y = -750409 , label = "US 54 & Others 85", 
           fontface = "italic", color = "black", size = 4) +
  coord_fixed() +
  scale_fill_continuous( type = "viridis",
    limits = c(0,900), breaks = c(25,50,75,100,250,500,750,900),
    guide = guide_colourbar(nbin = 10, draw.ulim = FALSE, draw.llim = FALSE, barwidth = 15,
                            title = "Travel volume (Daily)",
                            label.theme = element_text(angle = 90, size=7.8))
    ) +
  theme_mapcan() + theme(legend.position="", 
                         plot.title=element_text(size=rel(1.3), face="bold"))+
  ggtitle("During the pandemic")

m3 + m4 
```


##plots
```{r}
(p6+p5)/(m3+m4 ) + plot_layout(heights = unit(c(6.5, 1 ), c('cm', 'null'))) + plot_annotation(tag_levels = 'A')
ggsave("Figure/Figure1.png", width = 14, , height = 12, dpi= 500)


(p7)/(m1+m2 ) + plot_layout(heights = unit(c(5.8, 1 ), c('cm', 'null'))) + plot_annotation(tag_levels = 'A')
ggsave("Figure/Figure2.png", width = 9, , height = 10, dpi= 500)

(p1 + p2 ) + plot_annotation(tag_levels = 'A')  
ggsave("Figure/FigureApp2.png", width = 10, , height = 6, dpi =500)


# (p3 + p4) + plot_annotation(tag_levels = 'A')
# ggsave("Figure/diffsource.png", width = 10, , height = 6, dpi =1200)

```

## get Daily total volume (during the pandemic) for empi model (rest of manuscript)
```{r}
daily.TV <- function(){
  Daily1 <- total.pand(TDF, FC, FCEXCrew)
  Daily1 <- select(Daily1, c("date", "TDF_CA", "TDF_Int", "FC", "crew_CA", "Exp_CA" ))
  #Find exemption of TDF for INT and CA
  Daily1["exclINT"] <- Daily1$FC - Daily1$TDF_Int
  Daily1["exclCA"] <- Daily1$crew_CA + Daily1$Exp_CA
  #Share of each province during the pandeic
  perCAPan <- per12 %>% filter(per12$date >= as.Date("2020-09-01"))
  perCAPan <- select(perCAPan, -c(nonUS, US))
  ## divid exemption Canada based perCAPan between provinces per month
  Dailypro <- perCAPan
  for (i in 1:nrow(Dailypro)){
    for (j in 2:length(Dailypro)){
      Dailypro[i,j] <- (Daily1$exclCA[i])*(perCAPan[i,j]/100) }
  }
  ## make a dataframe of exemption Travel volume (who did not fill TDF form) for province and INT
  Dailypro <- Dailypro %>% left_join(select(Daily1, c(date, exclINT)), by ="date")
  # make data exemption daily 273
  Dailypro["Ym"] <- format(as.Date(Dailypro$date), "%Y-%m")
  Daily2 <- Dailypro %>% select(date) %>% 
    complete(date=  seq.Date(min(date), max(date)+30, by="day"),) %>%
    mutate(Ym=paste(year(date),str_pad(month(date), 2, pad = "0"),sep = "-"))
  Daily2 <- Daily2 %>% left_join(Dailypro, by="Ym")
  
  return(Daily2)
}

Daily2 <- daily.TV()

# divide number for days in a month
v_daily <- function(yr_month, v_monthly) {
   # Convert current month to current date
   curr_date <- ymd(paste0(yr_month, "-1"))
   # Calculate number of days in the month
   n_days <- days_in_month(curr_date)
  # Calculate daily rate
   (v_monthly) / (n_days)
}

Daily2 <- Daily2 %>%
   mutate(ON = map2_dbl(Ym, ON, v_daily), QC = map2_dbl(Ym, QC, v_daily), 
          MB = map2_dbl(Ym, MB, v_daily), NB = map2_dbl(Ym, NB, v_daily),
          AB = map2_dbl(Ym, AB, v_daily), NS = map2_dbl(Ym, NS, v_daily), 
          BC = map2_dbl(Ym, BC, v_daily), SK = map2_dbl(Ym, SK, v_daily), 
          PEI = map2_dbl(Ym, PEI, v_daily), TR = map2_dbl(Ym, TR, v_daily),
          exclINT = map2_dbl(Ym, exclINT, v_daily)
          )
 
#keep the interested columns
Daily2 <- select(Daily2, -c("date.y", "Ym")) %>% rename(date = date.x)
Daily2 <- Daily2 %>% rename(INT= exclINT)

## add daily exemption into raw TDF data source
TDF <- select(TDF, -c("US", "nonUS","ALLCANotNL")) %>% rename(INT = AllFR)
Tra_daily2 <- TDF
origin <- c("ON","QC","MB", "NB", "AB", "NS", "BC", "SK", "PEI", "TR","INT")
for (i in 1:length(origin)){
  pr <- origin[i]
  for (j in 1:nrow(Tra_daily2)){
    Tra_daily2[j, pr] <- TDF[j, pr] + Daily2[j,pr]
  }
}
# round data 
Tra_daily2 <- data.frame(lapply(Tra_daily2, function(x) if(is.numeric(x)) floor(x) else x))

#write.csv(Tra_daily2, " TotalTraVolume.csv")
```


